A spiral galaxy's mass distribution uncovered through lensing and dynamics

### 2015-11-13 docs/model_JAM_v3.01.tex
Included corrections in abstract and introduction.

\paragraph{Data comparison.} As data we use stellar line-of-sight rotation velocities $v_\text{rot} \equiv \langle v_\text{los} \rangle$ and velocity dispersions $\sigma$ as described in Section \ref{sec:data}. The JAM models give us a prediction for the second line-of-sight velocity moment $\langle v_\text{los}^2 \rangle$. The root mean square (rms) line-of-sight velocity $v_\text{rms}$ allows a data-model comparison by relating theses velocities according to 
\begin{equation*}
v_\text{rms}^2 = \langle v_\text{los}^2 \rangle = v_\text{rot}^2 + \sigma^2.
\end{equation*}
The model in Equation \ref{eq:explicitLOSvelCap} predicts the intrinsic $\langle v_\text{los}^2 \rangle$ at a given position on the sky, which have then to be modified according to the mode of observation, to be comparable to the measurements. The measured $v_\text{rms}$ is a light-weighted mean for a pixel along the long-slit of the spectrograph, with height $L_y = 1$ arcsec \citep{SWELLSV} and a certain given extent in along the galaxy's major axis, $L_x$, i.e. for a rectangular aperture
\begin{equation*}
\text{AP}(x',y') = \left\{ \begin{array}{ll} 1 & \text{for } -\frac{L_x}{2} \leq x' \leq + \frac{L_x}{2}\\
& \text{and } - \frac{L_y}{2} \leq y' \leq + \frac{L_y}{2}  \\ 0 & \text{otherwise.} \end{array} \right.
\end{equation*}
The light arriving at the spectrograph itself was subject to seeing, i.e. a Gaussian with Full Width Half Maximum (FWHM) of 1.1 arcsec \citep{SWELLSV}
\begin{equation*}
\text{PSF}(x',y')=\mathscr{N}(0,\text{FWHM}/2\sqrt{2\ln2}).
\end{equation*}
The model predictions have therefore to be convolved with the convolution kernel
\begin{eqnarray*}
K(x',y') &=& (\text{PSF} \ast \text{AP})(x',y') \\
&=& \frac{1}{4} \prod_{u \in \{x',y'\}} \left[ \text{erf}\left( \frac{L_u/2 - u}{\sqrt{2}\sigma_\text{seeing}}\right) + \text{erf} \left( \frac{L_u/2 + u}{\sqrt{2} \sigma_\text{seeing}} \right) \right]
\end{eqnarray*}
and weighted by the surface brightness $I(x',y')$
\begin{eqnarray*}
I_\text{obs} &=& I \ast K\\
\langle v_\text{los}^2 \rangle |_\text{obs} &=& \frac{(I \langle v_\text{los}^2\rangle) \ast K}{I_\text{obs}}.
\end{eqnarray*}
If provided with the convolution kernel, the JAM code\footnote{The IDL code package for Jeans Axisymmetric Modelling (JAM) by \citet{Cap08} is available online at \url{http://www-astro.physics.ox.ac.uk/~mxc/software}. The version from June 2012 was used in this work.} by \citet{Cap08} performs the convolution numerically. We set $L_x = 0.21$ arcsec as the width of the model pixel, and get a prediction for the actual measurements in bins of width 0.63, 1.26 and 1.89 arcsec \citep{SWELLSV} as light-weighted mean from each 3, 6 and 9 model pixels.

\paragraph{Rotation curve.} The intrinsic rotation curve is the first velocity moment $\langle v_\phi\rangle = \sqrt{\langle v_\phi^2 \rangle - \sigma_\phi^2}$. The observed rotation curve is the projection of the light-weighted contributions to $\langle v_\phi\rangle$ along the line-of-sight(\cite{Cap08}),
\begin{equation*}
I \langle v_\text{los}\rangle = \int_{-\infty}^{+\infty} \nu \langle v_\phi\rangle \cos \phi \sin i \diff z'.
\end{equation*}
The first velocity moments cannot be uniquely determined from the Jeans equations, which give only a prediction for the second velocity moments. Further assumptions are needed to separate the second velocity moments into ordered and random motion. \citet{Cap08} assumes that in a steady state there is no streaming velocity in $R$ direction, i.e. $\langle v_R \rangle = 0$ and therefore $\sigma_R^2 = \langle v_R^2 \rangle$. Then \citet{Cap08} relates the dispersions in $R$ and $\phi$ direction such that
$$\langle v_\phi\rangle = \sqrt{\langle v_\phi^2 \rangle - \sigma_\phi^2} \equiv \kappa \sqrt{\langle v_\phi^2 \rangle - \langle v_R^2 \rangle},$$
and the $\kappa$ parameter quantifies the rotation: $\kappa = 0$ means no rotation at all and $|\kappa| = 1$ describes a velocity dispersion ellipsoid that is a circle in the $R$-$\phi$ plane \citep{Cap08}. The sign of $\kappa$ determines the rotation direction. We can assign a constant $\kappa_k$ to every Gaussian in the MGE formalism and
\begin{equation*}
\nu \langle v_\phi\rangle = \left[\nu \sum_{k} \kappa_k^2 \left( [\nu\langle v_\phi^2 \rangle]_k - [\nu\langle v_R^2 \rangle]_k\right) \right]^{1/2}
\end{equation*} 
is then the light-weighted circular velocity curve, given the second velocity moments found from the Jeans equations.
\\To model the counter-rotating core of J1331 with one free parameter for, we employ the condition that the overall $\kappa(R)$ profile should smoothly and relatively steeply transition from $\kappa(R) = -\kappa' < 0$ at small $R$ through $\kappa(R_0) \equiv 0$ and increase to $\kappa(R) = \kappa' > 0$ at large $R$. Our imposed profile is
\begin{equation}
\kappa(R) = \kappa' \frac{R^2 - R_0^2}{R^2 + R_0^2}. \label{eq:kappa_profile}
\end{equation}
We find $\kappa'$ by matching the model $\langle v_\text{los} \rangle$ with the symmetrized $v_\text{rot}$ data, where for a given $\kappa'$ the $\kappa_k$ are found from fitting the MGE generated profile $\kappa(R) = \sum_k \kappa_k \nu_k(r)/\sum_k \nu_k(r)$ to Equation \ref{eq:kappa_profile}. The observed zero point is at $R'_0\approx 2$ arcsec. In the deprojected galactic plane the radius of zero rotation would be at a $R_0 \gtrsim 2$ arcsec, and we choose it to be at 2.2 arcsec.

>>>>

\paragraph{Data comparison.} As data we use stellar line-of-sight rotation velocities $v_\text{rot} \equiv \langle v_\text{los} \rangle$ and velocity dispersions $\sigma$ as described in Section \ref{sec:data}. The JAM models give us a prediction for the second line-of-sight velocity moment $\langle v_\text{los}^2 \rangle$. The root mean square (rms) line-of-sight velocity $v_\text{rms}$ allows a data-model comparison by relating theses velocities according to 
\begin{equation*}
v_\text{rms}^2 = \langle v_\text{los}^2 \rangle = v_\text{rot}^2 + \sigma^2.
\end{equation*}
The model in Equation \ref{eq:explicitLOSvelCap} predicts the intrinsic $\langle v_\text{los}^2 \rangle$ at a given position on the sky, which have then to be modified according to the mode of observation, to be comparable to the measurements. The measured $v_\text{rms}$ is a light-weighted mean for a pixel along the long-slit of the spectrograph, with height $L_y = 1''$ \citep{SWELLSV} and a certain given extent in along the galaxy's major axis, $L_x$, i.e. for a rectangular aperture
\begin{equation*}
\text{AP}(x',y') = \left\{ \begin{array}{ll} 1 & \text{for } -\frac{L_x}{2} \leq x' \leq + \frac{L_x}{2}\\
& \text{and } - \frac{L_y}{2} \leq y' \leq + \frac{L_y}{2}  \\ 0 & \text{otherwise.} \end{array} \right.
\end{equation*}
The light arriving at the spectrograph itself was subject to seeing, i.e. a Gaussian with Full Width Half Maximum (FWHM) of $1.1''$ \citep{SWELLSV}
\begin{equation*}
\text{PSF}(x',y')=\mathscr{N}(0,\text{FWHM}/2\sqrt{2\ln2}).
\end{equation*}
The model predictions have therefore to be convolved with the convolution kernel
\begin{eqnarray*}
K(x',y') &=& (\text{PSF} \ast \text{AP})(x',y') \\
&=& \frac{1}{4} \prod_{u \in \{x',y'\}} \left[ \text{erf}\left( \frac{L_u/2 - u}{\sqrt{2}\sigma_\text{seeing}}\right) + \text{erf} \left( \frac{L_u/2 + u}{\sqrt{2} \sigma_\text{seeing}} \right) \right]
\end{eqnarray*}
and weighted by the surface brightness $I(x',y')$
\begin{eqnarray*}
I_\text{obs} &=& I \ast K\\
\langle v_\text{los}^2 \rangle |_\text{obs} &=& \frac{(I \langle v_\text{los}^2\rangle) \ast K}{I_\text{obs}}.
\end{eqnarray*}
If provided with the convolution kernel, the JAM code\footnote{The IDL code package for Jeans Axisymmetric Modelling (JAM) by \citet{Cap08} is available online at \url{http://www-astro.physics.ox.ac.uk/~mxc/software}. The version from June 2012 was used in this work.} by \citet{Cap08} performs the convolution numerically. We set $L_x = 0.21''$ as the width of the model pixel, and get a prediction for the actual measurements in bins of width $0.63''$, $1.26''$ and $1.89''$ \citep{SWELLSV} as light-weighted mean from each 3, 6 and 9 model pixels.

\paragraph{Rotation curve.} The intrinsic rotation curve is the first velocity moment $\langle v_\phi\rangle = \sqrt{\langle v_\phi^2 \rangle - \sigma_\phi^2}$. The observed rotation curve is the projection of the light-weighted contributions to $\langle v_\phi\rangle$ along the line-of-sight(\cite{Cap08}),
\begin{equation*}
I \langle v_\text{los}\rangle = \int_{-\infty}^{+\infty} \nu \langle v_\phi\rangle \cos \phi \sin i \diff z'.
\end{equation*}
The first velocity moments cannot be uniquely determined from the Jeans equations, which give only a prediction for the second velocity moments. Further assumptions are needed to separate the second velocity moments into ordered and random motion. \citet{Cap08} assumes that in a steady state there is no streaming velocity in $R$ direction, i.e. $\langle v_R \rangle = 0$ and therefore $\sigma_R^2 = \langle v_R^2 \rangle$. Then \citet{Cap08} relates the dispersions in $R$ and $\phi$ direction such that
$$\langle v_\phi\rangle = \sqrt{\langle v_\phi^2 \rangle - \sigma_\phi^2} \equiv \kappa \sqrt{\langle v_\phi^2 \rangle - \langle v_R^2 \rangle},$$
and the $\kappa$ parameter quantifies the rotation: $\kappa = 0$ means no rotation at all and $|\kappa| = 1$ describes a velocity dispersion ellipsoid that is a circle in the $R$-$\phi$ plane \citep{Cap08}. The sign of $\kappa$ determines the rotation direction. We can assign a constant $\kappa_k$ to every Gaussian in the MGE formalism and
\begin{equation*}
\nu \langle v_\phi\rangle = \left[\nu \sum_{k} \kappa_k^2 \left( [\nu\langle v_\phi^2 \rangle]_k - [\nu\langle v_R^2 \rangle]_k\right) \right]^{1/2}
\end{equation*} 
is then the light-weighted circular velocity curve, given the second velocity moments found from the Jeans equations.
\\To model the counter-rotating core of J1331 with one free parameter for, we employ the condition that the overall $\kappa(R)$ profile should smoothly and relatively steeply transition from $\kappa(R) = -\kappa' < 0$ at small $R$ through $\kappa(R_0) \equiv 0$ and increase to $\kappa(R) = \kappa' > 0$ at large $R$. Our imposed profile is
\begin{equation}
\kappa(R) = \kappa' \frac{R^2 - R_0^2}{R^2 + R_0^2}. \label{eq:kappa_profile}
\end{equation}
We find $\kappa'$ by matching the model $\langle v_\text{los} \rangle$ with the symmetrized $v_\text{rot}$ data, where for a given $\kappa'$ the $\kappa_k$ are found from fitting the MGE generated profile $\kappa(R) = \sum_k \kappa_k \nu_k(r)/\sum_k \nu_k(r)$ to Equation \ref{eq:kappa_profile}. The observed zero point is at $R'_0\approx 2''$. In the deprojected galactic plane the radius of zero rotation would be at a $R_0 \gtrsim 2''$, and we choose it to be at $2.2''$.

### 2015-11-13 docs/model_MGE_v3.01.tex
Included corrections in abstract and introduction.

\subsection{Multi-Gaussian Expansion (MGE) formalism} \label{sec:MGE_theo}

Multi-Gaussian Expansions (MGE) are used to parametrize the observed surface brightness or projected total mass of a galaxy as a sum of $N$ two-dimensional, elliptical Gaussians \citep{1991ApJ...366..599B,1992A&A...253..366M,1994A&A...285..723E,1999MNRAS.303..495E}. This work makes use of the algorithm and code\footnote{The IDL code package for fitting MGEs to images by \citet{Cap02} is available online at \url{http://www-astro.physics.ox.ac.uk/~mxc/software}. The version from June 2012 was used in this work.} by \citet{Cap02}. We assume all Gaussians to have the same center and position angle $\phi$, i.e. orientation of the Gaussians' major axis measured counter-clockwise from the $y'$-axis of the coordinate system with polar coordinates $(R',\theta')$. Then the surface brightness can be written as
\begin{eqnarray}
I(R',\theta') &=& \sum_{i=1}^{N} I_{0,i} \exp\left[ - \frac{1}{2\sigma_i^2} \left(x^{'2} + \frac{y^{'2}}{q_i^{'2}}\right)\right]\label{eq:MGEgeneral}\\
\text{with } I_{0,i} &=& \frac{L_i}{2\pi \sigma_i^2 q'_i}\label{eq:centralItotalL}\\
\text{and } x'_i &=& R' \cos(\theta' - \phi)\nonumber\\
y'_i &=& R' \sin(\theta' - \phi),\nonumber
\end{eqnarray}
where $I_{0,i}$ is the central surface brightness of each Gaussian, $L_i$ its total luminosity, $\sigma_i$ its dispersion along the major axis and $q'_i$ the axis ratio between the elliptical Gaussians major and minor axis.
\\We can also expand the telescopes point-spread function (PSF) as a sum of circular Gaussians,
\begin{equation}
\text{PSF}(x',y') = \sum_j \frac{G_j}{2 \pi \delta_j^2} \exp\left[- \frac{1}{2 \delta_j^2} \left({x'}^2 + {y'}^2 \right)\right], \label{eq:PSFgeneral}
\end{equation}
where $\sum_j G_j = 1$ and $\delta_j$ are in this case the dispersions of the circular PSF Gaussians. In this case the observed surface brightness distribution is a convolution of the intrinsic surface brightness in Equation \ref{eq:MGEgeneral} with the PSF in Equation \ref{eq:PSFgeneral}: $(I \ast \text{PSF}) (x',y')$ is then again a sum of Gaussians and can be directly fitted to an image of the galaxy in question.

>>>>

\subsection{Surface brightness model} \label{sec:MGE_theo}

\paragraph{Multi-Gaussian Expansion (MGE).} MGEs are used to parametrize the observed surface brightness or projected total mass of a galaxy as a sum of $N$ two-dimensional, elliptical Gaussians \citep{1991ApJ...366..599B,1992A&A...253..366M,1994A&A...285..723E,1999MNRAS.303..495E}. This work makes use of the algorithm and code\footnote{The IDL code package for fitting MGEs to images by \citet{Cap02} is available online at \url{http://www-astro.physics.ox.ac.uk/~mxc/software}. The version from June 2012 was used in this work.} by \citet{Cap02}. We assume all Gaussians to have the same center and position angle $\phi$, i.e. orientation of the Gaussians' major axis measured counter-clockwise from the $y'$-axis of the coordinate system with polar coordinates $(R',\theta')$. Then the surface brightness can be written as
\begin{eqnarray}
I(R',\theta') &=& \sum_{i=1}^{N} I_{0,i} \exp\left[ - \frac{1}{2\sigma_i^2} \left(x^{'2} + \frac{y^{'2}}{q_i^{'2}}\right)\right]\label{eq:MGEgeneral}\\
\text{with } I_{0,i} &=& \frac{L_i}{2\pi \sigma_i^2 q'_i}\label{eq:centralItotalL}\\
\text{and } x'_i &=& R' \cos(\theta' - \phi)\nonumber\\
y'_i &=& R' \sin(\theta' - \phi),\nonumber
\end{eqnarray}
where $I_{0,i}$ is the central surface brightness of each Gaussian, $L_i$ its total luminosity, $\sigma_i$ its dispersion along the major axis and $q'_i$ the axis ratio between the elliptical Gaussians major and minor axis.
\\We can also expand the telescopes point-spread function (PSF) as a sum of circular Gaussians,
\begin{equation}
\text{PSF}(x',y') = \sum_j \frac{G_j}{2 \pi \delta_j^2} \exp\left[- \frac{1}{2 \delta_j^2} \left({x'}^2 + {y'}^2 \right)\right], \label{eq:PSFgeneral}
\end{equation}
where $\sum_j G_j = 1$ and $\delta_j$ are in this case the dispersions of the circular PSF Gaussians. In this case the observed surface brightness distribution is a convolution of the intrinsic surface brightness in Equation \ref{eq:MGEgeneral} with the PSF in Equation \ref{eq:PSFgeneral}: $(I \ast \text{PSF}) (x',y')$ is then again a sum of Gaussians and can be directly fitted to an image of the galaxy in question.
